---
title: "Tissue-aware normalization"
author: "Chi-Lam Poon"
date: "12/30/2019"
output: html_document
---


## Normalization methods

We aimed to compare some commonly seen methods and see their effects on normalization. Their introductions can see this post: [Normalization methods for RNA-seq](https://chilampoon.github.io/2019/12/17/RNA-norm/)

- TMM
- RLE
- UQ
- DESeq
- QN
- qsmooth
- none

```{r setup} 
suppressPackageStartupMessages({
  library(yarn)
  library(stringr)
  library(Biobase)
  library(dplyr)
  library(rafalib)
  library(readr)
  library(ggplot2)
  library(ggpubr)
  library(edgeR)
})

source('~/cvd/v8/scripts/tissueAwareNorm.R')
```

## Load data
```{r}
eset <- readRDS('~/cvd/v8/exprs/gtex.filted.rds')
```


## Comparison of normalization methods

```{r comp}
# select some tissues
selectt <- c('Brain - Amygdala', 'Artery - Aorta', 'Heart - Atrial Appendage', 'Whole Blood', 'Testis')
seset <- eset[, which(pData(eset)$SMTSD %in% selectt)]
dim(seset)

dist.list <- list()
methods <- c('TMM', 'RLE', 'upperquartile', 'DESeq', 'QN', 'qsmooth', 'none')
for (method in methods) {
  dist.list[[method]] <- tissue.aware.norm(seset, "SMTSD", method)
}
#saveRDS(dist.list, file='~/cvd/v8/exprs/norm.cpm.rds')
```


Visualization. Note that those densities are log2 (count + 1), not cpm!
```{r vis, fig.height=13, fig.width=13}
## auxilary functions
getColors <- function(vec) {
  # assgin a color to each group of samples
  library(RColorBrewer)
  n <- length(unique(vec))
  col <- brewer.pal(n, "Paired")
  #col <- brewer.pal.info[brewer.pal.info$category=='qual', ] # get max. 74 colours
  #col_all <- unlist(mapply(brewer.pal, col$maxcolors, rownames(col)))
  ifelse (n > length(col), 
          cvec <- sample(col, n, replace=T),
          cvec <- sample(col, n, replace=F)
  )
  vec <- as.character(vec)
  names(vec) <- rep(NA, length(vec))
  for (g in 1:length(unique(vec))) {
    names(vec)[which(vec==unique(vec)[g])] <- cvec[g]
  }
  vec
}


dens.plot <- function(table, colors, ...) {
  cols <- names(colors)
  d <- plot(density(table[, 1]), col=cols[1], 
            lwd=2, las=2, xlab="", ...) +
    abline(v=0, lty=3) + title(xlab="log2 exprs", ylab=NA) +
    for (i in 2:ncol(table)) {
      den <- density(table[, i])
      lines(den$x, den$y, col=cols[i], lwd=2)
    } 
  legend('topright', legend=unique(colors), lty=1, col=unique(names(colors)), cex=0.6)
  d
}


sub.p <- pData(seset)
sub.p$SMTSD <- factor(sub.p$SMTSD)
sub.p <- sub.p[order(sub.p$SMTSD),]
colors <- getColors(sub.p$SMTSD)

mypar(3, 3)
for (method in methods) {
  print(dens.plot(dist.list[[method]][,match(rownames(sub.p), colnames(dist.list[[method]]))], colors, ylim=c(0, 0.3), main=method))
}
```

We chose to use __tissue-aware quantile normalization__.


## Data normalization
Quantile normalize each tissue in a tissue-aware manner.
```{r norm, eval=T}
storageMode(eset) <- "environment"
normalizedMatrix <- tissue.aware.norm(eset, "SMTSD", 'QN')
all(colnames(normalizedMatrix)==rownames(pData(eset)))
assayData(eset)[["normalizedMatrix"]] <- normalizedMatrix
storageMode(eset) <- "lockedEnvironment"
saveRDS(eset, file='~/cvd/v8/exprs/gtex.filted.norm.rds')
```


### Session info
```{r session}
sessionInfo()
```

