---
title: "limma-voom MDD"
author: "Chi-Lam Poon"
date: "10/18/2020"
output: html_document
---



According to paper [Batch correction evaluation framework using a-priori gene-gene associations: applied to the GTEx dataset](https://bmcbioinformatics.biomedcentral.com/articles/10.1186/s12859-019-2855-9), the linear regression correction method combining ComBat outperformed other methods. We chose to use their linear model including:

- Gender (GENDER)
- Ischemic time (SMTSISCH)
- Age (AGE)
- Experimental batch (SMGEBTCH)
- Death type (DTHHRDY)

Tissues:

$$Y \sim \beta_1GENDER + \beta_2SMTSISCH + \beta_3AGE + \beta_4SMNABTCHT + \beta_5DTHHRDY + \beta_6MHCVD + \epsilon$$




```{r setup}
suppressPackageStartupMessages({
  library(purrr)
  library(dplyr)
  library(metap)
  library(knitr)
  library(ggplot2)
  library(ggpubr)
  library(Biobase)
})

source("~/cvd/v8/scripts/utils.R")
```


## Load data
```{r load, eval=T}
eset <- readRDS('~/cvd/v8/exprs/gtex.mdd.filted.norm.rds')
dim(eset)
```



## Limma-voom
### Design matrix without RACE
```{r}
table_dir_1 <- '~/cvd/v8/results/new/mdd/tables'
plot_dir_1 <- '~/cvd/v8/results/new/mdd/plots'

### not merge!
tissues <- as.character(unique(pData(eset)$SMTSD))
disease <- "mdd1"
# sex-specific tissues
oneSexTissues <- c('Testis', 'Ovary', 'Uterus', 'Vagina', 'Prostate')
# tissues with only one batch
num <- sapply(tissues, function(x) length(unique(pData(eset)[pData(eset)$SMTSD == x,]$SMNABTCHT)))
oneBTCHtissues <- names(num[num==1])

for (t in tissues) {
  try({
    print(t)
    # Subset phenotype and expression data
    sub_eset <- eset[,which(pData(eset)$SMTSD %in% t)]
    sub_pdata <- pData(sub_eset)
    sub_exprs <- assayData(sub_eset)[['normalizedMatrix']]
    
    # Construct the linear model
    print("Running limma pipeline...")
    age    <- as.numeric(sub_pdata$AGE)
    sex    <- factor(sub_pdata$SEX)
    batch  <- factor(sub_pdata$SMNABTCHT)
    mdd    <- factor(sub_pdata$MHDPRSSN)
    subject<- factor(sub_pdata$SUBJID)
    hardy  <- factor(as.character(sub_pdata$DTHHRDY))
    ischtime <- as.numeric(sub_pdata$SMTSISCH)
  
    
    if (t %in% oneSexTissues) {
      design <- model.matrix(~ age + batch + mdd + hardy + ischtime)
    } else if (t %in% oneBTCHtissues) {
      design <- model.matrix(~ sex + age + mdd + hardy + ischtime)
    } else {
      design <- model.matrix(~ sex + age + batch + mdd + hardy + ischtime)
    }
  
    
    # Use voom weighting
    transformedCounts <- sub_exprs
    voomOutput <- voomWeightsCustomized(transformedCounts, design)
  
    fit <- lmFit(voomOutput, design)
    fit <- eBayes(fit)
    gl <- fData(eset)$geneName
    topTable <- topTable(fit, number=Inf, genelist=gl, coef=disease)
    
    # Output reports & plots
    write.table(topTable, file=sprintf("%s/%s.txt", table_dir_1, t), sep="\t", quote=F)
    pdf(sprintf("%s/%s.volcanoplot.pdf", plot_dir_1, t))
      limma::volcanoplot(fit, coef=disease, highlight = 6, names = gl, main = t)
    dev.off()
    pdf(sprintf("%s/%s.MAplot.pdf", plot_dir_1, t))
      limma::plotMA(fit, main = t, coef=disease)
    dev.off()
  })
}

```




